---
title: "HW 12: Population Scale Analysis"
author: "Amanda Wilpitz | A17463962"
format: pdf
---

> Q13: Read this file into R and determine the sample size for each geneotype and their corresponding median expression levels for each of these genotypes.

```{r}
data <- read.table("rs8067378_ENSG00000172057.6.txt")
head(data)
```

```{r}
table(data$geno)
```
```{r}
lapply(split(data$exp, data$geno), summary)
```

> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

```{r}
library(ggplot2)
```

```{r}
ggplot(data) + 
  aes(geno, exp, fill = geno) +
  geom_boxplot(notch=T) +
  labs(x = "Genotype", y = "Expression")
  
```

I can infer that A/A has higher expression levels and has a difference of about 10 between A/A and G/G.

Since the expression levels differ across the genotypes enough to visually see, the SNP could effect the expression of ORMDL3.
